Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-05-29 manuscript.Rmd
Edits to Method section

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated
by quantifying the frequency of shared content words with respect to those
words’ frequencies in the entire tweet corpus. If a linguistic term has a
high frequency in one tweet but has a low frequency in the whole tweet
corpus, then it is more likely a content word and is this given more weight
in calculating social cohesion.

The tweets were first sorted into ascending order by timestamp. They were
then grouped into successive sets of five tweets. Inverse frequency
weighting for the overlap words and all words was conducted for each pair
of two tweets in the group of five, then divided to get the social cohesion
score for that pair (see Eq. 1-3). This resulted in ten social cohesion
values for each group of five tweets, which were then averaged. Finally,
the mean of all average cohesion values for all groups was taken to get
a social cohesion score for each day.

@@

The social cohesion calculation was performed on a subset of the corpus
containing (recalculate with new date span) English and (recalculate with
new date span) Arabic tweets. All hashtags and URLs were included in the
analysis. This method is similar to that of inverse document frequency (IDF).
For more information on IDF, see @jones1972statistical and
@robertson2004understanding.

### ICEWS Event Time Series
Using the Integrated Crisis Early Warning System (ICEWS) developed at the
Defense Advanced Research Project Agency, the daily events in Syria were
scored with an intensity rating ranging from -10 to 10. Negative events,
such as riots and attacks, are coded with low negative numbers while
positive, peaceful events are coded with high positive numbers. This
well-validated coding scheme allows for numerical quantification of
different event types.

For both the target only and the target and source data, time series of
the count of all events were generated by summing the total number of events
on each day, irrespective of ICEWS score. Similarly, time series of count
of positive events were generated by summing the number of ICEWS events
for each day that had an intensity rating greater than zero. Lastly,
time series of count of negative events were generated by summing the number
of ICEWS events for each day that had an intensity rating less than zero.

## Analyses
### Cross-Recurrence Quantification Analysis
We first used cross-recurrence quantification analysis
[CRQA; @zbilut1998detecting] to investigate the temporal patterns of
social cohesion and real-world events in Syria during Arab Spring. As
an extension of RQA, which captures the patterns of revisited states of
a single time series [@zbilut1992embeddings; @webber1994dynamical], CRQA
allows for two time series to be projected onto one another to investigate
shared states and trajectories. Nonlinear analyses such as CRQA can reveal
dynamics that are not captured by traditional linear methods.

In order to perform CRQA, the time series need to be converted to the same
scale, while still preserving a sampling rate that respects the native
timescale of the system
[cf. Nyquist frequency in signal processing; @grenander1959probability].
We converted each time series (e.g., Figure \@ref(fig:raw-ts)) into deciles
(e.g., Figure \@ref(fig:deciled-ts)) to analyze the dynamics of different
levels of social cohesion and counts of all, positive, and negative events
on each day. In CRQA, a recurrent point is when two signals share the same
state. These shared states can be at the same time or at later points in
the time series.

#### Visualization
A recurrence plot (RP) is the visual representation of CRQA. To construct
a recurrence plot, two time series are plotted against each other on the
X and Y axes, and at every point where they share the same state, and point
in placed on the plot to indicate recurrence. Visually analyzing these plots
can help reveal patterns of structure and recurrence. The line y = x is known
as the line of identity. This line falls where the time series match
temporally, and a filled line would indicate perfect synchrony between the
two time series. See
Figures \@ref(fig:plot-rp-targ-all)-\@ref(fig:plot-rp-source-targ-neg)
for examples of RPs.

#### Metrics
The percentage of recurrent points to total possible points is a metric
called recurrence rate (RR). This metrics gives a general description of
how frequently the time series are sharing the same states. In the current
study, a recurrent point would indicate that the level of social cohesion
and frequency of events were at the same intensity.

Recurrent points that occur in successive moments form line structures.
These are conceptualized as shared trajectories, where the two time series
are moving together across time in a shared state. The percentage of
recurrent points that make up diagonal line structures (i.e., two or more
consecutive points) is a metric called percent determinism (DET). For the
current project, a high DET would indicate that not only are the levels of
social cohesion and frequency of events occurring at similar intensities,
but they are doing so across consecutive days.

The average length of the line structures is called mean line (L). This can
be conceptualized as the average amount of time that the two time series move
together before one or both shift states. The longest line length is called
max line (maxL). maxL indicates the longest instance of being “stuck” in a
particular state together. The total number of lines is called NRLINE. The
more lines there are, the more coordinated the system may be. The
complexity or variability of line lengths is captured by a metric called
entropy (ENTR) and entropy while accounting for the total number of lines in
the plot is normalized entropy (rENTR). These entropy metrics help indicate
the stability and predictability of the system. When one of the signals gets
“stuck” in a state and the other moves on to other states, this results in
vertical line structures. The proportion of points forming vertical line
structures is a metric called laminarity (LAM) and the average length of the
vertical trajectories is called trapping time (TT). For more detail on CRQA
metrics, see @R-crqa.

#### Parameters
To conduct the analyses on the decile data, the parameters were set to those
typically used in categorical CRQA. Delay was set to zero, embedding dimension
was set to 1, and radius was set to .001 to allow for only exact recurrent
matches. No normalization was done, and lines were considered two or more
consecutive recurrent points. The Theiler window was set to zero to include
points along the line of identity. For more details on the CRQA parameters and
how they impact the analysis, see @R-crqa.

#### Permutation Test
Using the deciled, real time series of social cohesion, count of all events,
count of positive events, and count of negative events, a permutation test was
done to determine statistical difference from chance for each CRQA metric.
Using the sample function in base R [@R-base], 1000 permuted time series were
created for each of the real time series. A permutation test is similar to
bootstrapping, except that the sampling of observations is done without
replacement. CRQA was then run 1000 times for social cohesion and count of all
events, social cohesion and count of positive events, and social cohesion and
count of negative events using the permuted time series. This resulted in
1000 results for each of the CRQA metrics for each combination of time series.
Statistical significance from chance was set at the 95th and 99th percentiles
of the metrics from the permuted CRQ analyses. Given that a permutation test
is similar to bootstrapping but does not include replacement, all points were
included in each CRQA analysis, leaving RR the same across all analyses. Thus,
RR is not included in the inferential tests. For more information on permutation
tests, see @good2013permutation.

### Windowed Cross-Recurrence Quantification Analysis
We used windowed cross-recurrence quantification analysis
[WCRQA; @webber2005recurrence] to further investigate the evolution of the
temporal dynamics of social cohesion and real-world events. In this analysis,
a smaller “window” of time points is used to calculate CRQA and capture more
fine-grain dynamics of two time series. The window remains the same size and
“slides” diagonally up the line of identity. The changes in the CRQA metrics
across the windows can reveal changes in dynamics that can then be linked to
external, real-world events.

#### Metrics
Conducting WCRQA results in CRQA metrics (e.g., RR, DET, ENTR) for each
respective window. These metrics can then be plotted as a time series to
visualize their evolution over time. For these analyses, we decided to
visualize recurrence rate and determinism.

#### Parameters
In addition to the CRQA parameters, two other parameters are required when
conducting WCRQA. First, a window size must be decided on. The window size
will determine how many observations are used in each CRQA analysis. We
conducted WCRQA with a window size of 14 to investigate the dynamics over
two weeks and allow for the capture of more structure and thus a more
meaningful interpretation of DET. The second parameter to be chosen is a
step size. Window step size is how far the window moves up the line of identity
before conducting CRQA again. We chose 1 for the step size to understand
how the metrics evolve from day to day. All other CRQA metrics remained the
same as previously noted.

@@

#### Permutation Test
To visualize the differences in the WCRQA metrics from those obtained by
chance, a permutation test was done to set upper and lower bounds for the
95th and 99th percentiles. Using a sample size of 14, one-thousand permuted
time series were generated without replacement for each real time series. The
upper and lower 95th and 99th percentiles were set based on the distribution
of CRQA metrics from these one-thousand CRQ analyses.

>>>>

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated
by quantifying the frequency of shared content words with respect to those
words’ frequencies in the entire tweet corpus. If a linguistic term has a
high frequency in one tweet but has a low frequency in the whole tweet
corpus, then it is more likely a content word and is thus given more weight
in calculating social cohesion. 
This method is similar to inverse document frequency (IDF).
For more information on IDF, see @jones1972statistical and
@robertson2004understanding.

The tweets were first sorted into ascending order by timestamp. They were
then grouped into successive sets of five tweets. Inverse frequency
weighting for the overlap words and all words was conducted for each pair
of two tweets in the group of five, then divided to get the social cohesion
score for that pair (see Eq. 1-3). This resulted in ten social cohesion
values for each group of five tweets; these values were then averaged to get
a social cohesion score for the entire set of five tweets. Finally,
the mean of all average cohesion values for all groups was taken to get
a social cohesion score for each day.

@@

The social cohesion calculation was performed on the tweet corpus
containing (recalculate with new date span) English and (recalculate with
new date span) Arabic tweets. All hashtags and URLs were included in the
analysis.

### ICEWS Event Time Series
Using the Integrated Crisis Early Warning System (ICEWS) developed at the
Defense Advanced Research Project Agency, the daily events were
scored with an intensity rating ranging from -10 to 10. Negative 
events---such as riots and attacks---are coded with low negative numbers, while
positive or peaceful events are coded with high positive numbers. This
well-validated coding scheme allows for numerical quantification of
different event types.

For both the target only and the target and source data described above, time series of
the count of all events were generated by summing the total number of events
on each day, irrespective of ICEWS score. Similarly, time series of count
of positive events were generated by summing the number of ICEWS events
for each day that had an intensity rating greater than zero. Lastly,
time series of count of negative events were generated by summing the number
of ICEWS events for each day that had an intensity rating less than zero.

## Analyses
### Cross-Recurrence Quantification Analysis
We first used cross-recurrence quantification analysis
[CRQA; @zbilut1998detecting] to investigate the temporal patterns of
social cohesion and real-world events in Syria during Arab Spring. CRQA
is an extension of RQA, which captures the patterns of revisited states of
a single time series [@zbilut1992embeddings; @webber1994dynamical]. CRQA
allows for two time series to be projected onto one another to investigate
shared states and trajectories.
In CRQA, we track these shared states by identifying *recurrent points*,
or intersections of time during which two signals shared or revisited the same
state. These shared states can be at the same time or at later points in
the time series. We can further chart shared trajectories by examining
the occurrences and patterns of multiple sequential recurrent points.

To perform CRQA, the time series must be converted to the same
scale, while still preserving a sampling rate that respects the native
timescale of the system
[cf. Nyquist frequency in signal processing; @grenander1959probability].
Because the social cohesion and event data have different scales and values,
we converted each time series (e.g., Figure \@ref(fig:raw-ts)) into deciles
(e.g., Figure \@ref(fig:deciled-ts)) to analyze the relative dynamics of different
levels of social cohesion and counts of all, positive, and negative events
on each day.

#### Visualization
A cross-recurrence plot (CRP) is the visual representation of CRQA. To construct
a cross-recurrence plot, two time series are plotted against each other on the
$x$ and $y$ axes. At every point where they share the same state, a point
in placed on the plot to indicate recurrence. (For categorical CRQA, this process
is similar to the Cartesian product.)

We can visually analyze these plots
to reveal patterns of structure and recurrence. The line $y = x$ is known
as the *line of identity*. This line falls where the time series match
temporally. A filled line would indicate perfect in-phase synchrony between the
two time series. See
Figures \@ref(fig:plot-rp-targ-all)-\@ref(fig:plot-rp-source-targ-neg)
for examples of RPs.

#### Metrics
While visual inspections can be informative, we can derive a variety of 
quantitative metrics from the cross-recurrence plot. Each of these metrics
provides information into the pattern and dynamics of the new shared system
between the original time series. These metrics are also critical for inferential
statistics. We will focus most directly on two of these metrics: 
*recurrence rate* and *determinism*.

The percentage of recurrent points to total possible points is a metric
called the *recurrence rate* (*RR*). This metrics gives a general description of
how frequently the time series are sharing the same states. In the current
study, a recurrent point would indicate that the level of social cohesion
and frequency of events were at the same intensity.

Recurrent points that occur in successive moments form line structures.
These are conceptualized as shared trajectories, where the two time series
are moving together across time in a shared state. The percentage of
recurrent points on the plot that fall along diagonal line structures 
(i.e., two or more
consecutive points) is a metric called *percent determinism* 
(or simply *determinism*; *DET*). For the
current project, a high DET would indicate that not only are the levels of
social cohesion and frequency of events occurring at similar intensities,
but they are doing so across consecutive days.

A number of other metrics provide insight not only into the amount of structure
between the systems but about the nature of this shared structure.
The average length of the line structures is called *mean line* (L), which can
be conceptualized as the average amount of time that the two time series move
together before one or both shift states. The longest line length is called
*max line* (*maxL*); it indicates the longest instance of being “stuck” in a
particular state together. The total *number of lines* on the plot (*NRLINE*) 
scales with structure: The more lines that exist, the more coupled the systems are. The
complexity or variability of line lengths is captured by a metric called
*entropy* (*ENTR*)---which may also be normalized by the total number of lines
(i.e., *normalized entropy*; *rENTR*)---and indicates
the stability and predictability of the system. When one of the signals gets
“stuck” in a state and the other moves on to other states, this results in
vertical line structures, which are quantified in *laminarity* (*LAM*;
i.e., the proportion of points forming vertical line structures) 
and trapping time (*TT*; i.e., the average length of the
vertical trajectories). For more detail on CRQA metrics, see @R-crqa.

#### Parameters
To conduct the analyses on the decile data, the parameters were set to those
typically used in categorical CRQA: Delay was set to $0$, embedding dimension
was set to $1$, and radius was set to $.001$ to allow for only exact recurrent
matches. No normalization was done, and lines were considered two or more
consecutive recurrent points. The Theiler window was set to $0$ to include
points along the line of identity. For more details on the CRQA parameters and
how they impact the analysis, see @R-crqa.

#### Permutation Test
Unlike some nonlinear analyses (e.g., fractal analyses), CRQA metrics do not have
inherently meaningful values. As a result, CRQA is best used as a relative
metric---that is, comparing the CRQA values across conditions or subsets
of the data. Typically, meaningful differences between values are determined with
traditional inferential statistics. While this may not be a problem for 
experimentally derived datasets, it can pose a barrier to researchers using
small-$n$ data or conducting data-informed case studies.

For CRQA, this can be addressed using *permutation tests*.
Similar to more established methods of statistical significance in CRQA research
[which create surrogate time series; e.g., @paxton2017interpersonal],
the permutation test approach to significance testing allows the researcher
to use a time series as its own baseline by testing whether the *structure*
of the unfolding events of the time series---rather than the *raw frequency* of the 
events within the time series---cohere together more than would be expected
by chance. A permutation test is similar to
bootstrapping, except that the sampling of observations is done without
replacement. In this approach, we simply create a large number of permutations
of the original time series (i.e., breaking up the dependency across time points but
preserving the frequencies of the events), conduct CRQA for each of those
permutations, and then examine whether the real time series' CRQA metrics
exceed those that we would expect to see by chance. The proportion of times that
the real time series' values exceeds the surrogate time series' values can
be interpreted as the alpha criterion to determine significance. 
For more information on permutation
tests, see @good2013permutation.

Using the real deciles of social cohesion, count of all events,
count of positive events, and count of negative events, we conducted a series
of permutation tests 
(one for the comparison between social cohesion and each type of event count)
to determine statistical difference from chance for each CRQA metric.
Using the `sample` function in base R [@R-base], $1000$ permuted time series were
created for each of the real time series.  
CRQA was then run $1000$ times for social cohesion and count of all
events, social cohesion and count of positive events, and social cohesion and
count of negative events using the permuted time series. This resulted in
$1000$ results for each of the CRQA metrics for each combination of time series.
Statistical significance from chance was set at the 95th and 99th percentiles
of the metrics from the permuted CRQ analyses. Given that a permutation test
is similar to bootstrapping but does not include replacement, all points were
included in each CRQA analysis, leaving RR constant across all permutations. Thus,
RR is not included in the inferential tests.

### Windowed Cross-Recurrence Quantification Analysis
We used windowed cross-recurrence quantification analysis
[WCRQA; @webber2005recurrence] to further investigate the evolution of the
temporal dynamics of social cohesion and real-world events. In this analysis,
a smaller “window” of time points is used to calculate CRQA and capture more
fine-grained dynamics of two time series. The window remains the same size and
“slides” diagonally up the line of identity. The changes in the CRQA metrics
across the windows can reveal changes in dynamics that can then be linked to
external, real-world events.

#### Metrics
Conducting WCRQA results in CRQA metrics (e.g., RR, DET, ENTR) for each
respective window. These metrics can then be plotted as a time series to
visualize their evolution over time. For these analyses, we decided to
focus on recurrence rate and determinism.

#### Parameters
In addition to the CRQA parameters, two other parameters are required when
conducting WCRQA. First, the *window size*
will determine how many observations are used in each CRQA analysis. We
conducted WCRQA with a window size of $14$ to investigate the dynamics over
two weeks and allow for the capture of more structure and thus a more
meaningful interpretation of DET.

The second parameter to be chosen is a *window step size*. 
Window step size is how far the window moves up the line of identity
before conducting CRQA again. We chose $1$ for the step size to understand
how the metrics evolve from day to day.

All other CRQA metrics remained the same as those previously noted.

@@

#### Permutation Test
To visualize the differences in the WCRQA metrics from those obtained by
chance, a permutation test was done to set upper and lower bounds for the
95th and 99th percentiles. Using a sample size of 14, $1000$ permuted
time series were generated without replacement for each real time series. The
upper and lower 95th and 99th percentiles were set based on the distribution
of the permuted WCRQA metrics.
